---
title: "Actual causation"
author: "Andi Fugard (@andi@sciences.social)"
date: "23 Sept 2023"
output:
  html_document:
    df_print: paged
    toc: true
  html_notebook:
    code_folding: none
---

This R code implements ideas in Joseph Halpern's *Actual Causality*.


## Library setup

```{r}
library(conflicted)
library(tidyverse)
library(R6)
```

```{r}
conflicts_prefer(dplyr::filter)
```

## The R6 class that does all the magic

Comments coming on this one. {R6} is a lovely package; however, I'm not sure how best to use it in Markdown.

```{r CausalMod}
CausalMod <- R6Class("CausalMod",
  public = list(
    
    initialize = function(funs = NA, ranges = NA) {
      self$setFuns(funs)
      self$setRanges(ranges)
    },
    
    setFuns = function(funs) {
      private$funs <- funs
    },
    
    setRanges = function(ranges) {
      private$ranges <- ranges
    },
    
    update = function(var, newVal) {
      if (!var %in% self$getVars())
        stop(paste(var, "is not a variable"))
      if (!newVal %in% self$getRanges()[[var]])
        stop(paste(newVal, "is not in range",
                   paste(self$getRanges()[[var]], collapse = ", " )))
      
      private$funs[[var]] <- \() newVal
    },
    
    updateList = function(valList) {
      vars <- names(valList)
      
      for (v in vars) {
        self$update(v, valList[[v]])
      }
    },
    
    getFuns = function() {
      private$funs
    },
    
    getRanges = function() {
      private$ranges
    },
    
    getVars = function() {
      names(private$funs)
    },
    
    getState = function() {
      res <- as.list(rep(NA, length(private$funs)))
      names(res) <- names(private$funs)
  
      for (v in names(res)) {
        theArgs <- formalArgs(private$funs[[v]])
        theParams <- list()
        if (!is.null(theArgs)) {
          theParams <- res[theArgs]
        }
        res[[v]] <- do.call(private$funs[[v]], theParams)
      }

      res      
    },
    
    testProp = function(prop) {
      state <- self$getState()[(formalArgs(prop))]
      do.call(prop, state)
    },
    
    AC1 = function(XsProp, phiProp) {
      self$testProp(XsProp) && self$testProp(phiProp)
    },
    
    AC2 = function(XsProp, phiProp) {
      Xs <- formalArgs(XsProp)
      partitions <- private$listZWs(Xs, names(private$funs))
      res <- data.frame()
      
      for (Ws in partitions$Ws) {
        possibilities <- expand.grid(private$ranges[c(Xs, Ws)])
        Zs <- setdiff(names(private$funs), c(Xs, Ws))
        
        cat("\nTrying partition:\n")
        private$printPartition(Xs, Ws, Zs)

        for (r in seq_len(nrow(possibilities))) {
          cloned <- self$clone()
          thisWorld <- possibilities[r,,drop = FALSE]
          cloned$updateList(thisWorld |> as.list())
          if (!cloned$testProp(XsProp) && !cloned$testProp(phiProp)) {
            cat("AC2u(a) PASSED on: \n")
            cat("  ")
            private$printFlatList(thisWorld |> as.list())
            cat("\n")
            
            X_setting <- self$getState()[Xs]
            Z_setting <- self$getState()[Zs]
            W_setting <- cloned$getState()[Ws]
            ZW_setting <- c(Z_setting, W_setting)
            
            test_subs <- private$allSubsets(ZW_setting)
            test_subs <- lapply(test_subs, \(x) c(X_setting, x))

            cat("Testing AC2u(b) on subsets:\n")
            
            AC2b_res <- c()
            for (x in seq_len(length(test_subs))) {
              cat("  ")
              private$printFlatList(test_subs[[x]])
              
              testWorld <- self$clone()
              testWorld$updateList(test_subs[[x]])
              testResult <- testWorld$testProp(phiProp)
              
              if (testResult)
                cat("✓\n")
              else
                cat("✗\n") 
              
              AC2b_res <- c(AC2b_res, testResult)
            }

            if (all(AC2b_res)) {
              res <- bind_rows(res, thisWorld)
              cat("✓✓ AC2(a) and AC2u(b)!\n")
            }
            else {
              cat("AC2u(b) failed!\n")              
            }
          }
        }
      }
      cat("\n")
      res
    },
    
    AC = function(XsProp, phiProp) {
      c(AC1 = self$AC1(XsProp, phiProp),
        AC2u = (nrow(self$AC2(XsProp, phiProp)) > 0)
      )
    }
    
  ),
  
  private = list(
    funs   = list(),
    ranges = list(),
    
    allZWs = function(Xs, vars) {
      varsList <- as.list(rep(NA, length(vars)))
      names(varsList) <- vars
  
      for (v in vars) {
        if (v %in% Xs)
          varsList[[v]] <- c("Z")
        else
          varsList[[v]] <- c("Z", "W")
      }
  
      expand.grid(varsList, stringsAsFactors = FALSE)
    },
    
    listZWs = function(Xs, vars) {
      dat <- private$allZWs(Xs, vars)
      Zs <- list(rep(NA, nrow(dat)))
      Ws <- list(rep(NA, nrow(dat)))
  
      for (i in seq_len(nrow(dat))) {
        Zs[[i]] <- names(dat)[dat[i,,drop = FALSE] == "Z"]
        Ws[[i]] <- names(dat)[dat[i,,drop = FALSE] == "W"]
      }
    
      list(Ws = Ws, Zs = Zs)
    },
    
    allSubsets = function(theList) {
      vars <- names(theList)
      varsList <- as.list(rep(NA, length(vars)))
      names(varsList) <- vars
      
      for (v in vars) {
          varsList[[v]] <- c(TRUE, FALSE)
      }
      
      truthtab <- expand.grid(varsList, stringsAsFactors = FALSE)
      
      res <- list()
      for (r in seq_len(nrow(truthtab))) {
        vec <- truthtab[r,,drop = FALSE] |> t() |> as.vector()
        res <- c(res, list(theList[vec]))
      }
      
      res
    },
    
    printFlatList = function(flat) {
      if (length(flat) > 0)
        stopifnot(!is.null(names(flat)))
      
      cat("[ ")
      for (v in names(flat)) {
        cat(v)
        cat("←")
        cat(flat[[v]])
        cat(" ")
      }
      cat("]")
    },
    
    printListofLists = function(lists) {
      for (e in lists) {
        cat("  ")
        private$printFlatList(e)
        cat("\n")
      }
    },
    
    printPartition = function(Xs, Ws, Zs) {
      cat("  X = {")
      cat(paste(Xs, collapse = ", "))
      cat("}\n")
      cat("  W = {")
      cat(paste(Ws, collapse = ", "))
      cat("}\n")
      cat("  Z-X = {")
      cat(paste(Zs, collapse = ", "))
      cat("}\n")
    }
  )
)
```


## Easy example: one person lobbing stone at bottle


Here's an easy simplfication of an example given by Halpert. Suzy throws (ST), bottle smashes (BS).

```{r simpBottle}
simpBottle <- CausalMod$new(
  funs = list(
    ST = \() 1,
    BS = \(ST) ST
  ), 
  ranges = list(
    ST = c(0, 1),
    BS = c(0, 1)    
  )
)

simpBottle$AC(\(ST) ST == 1, \(BS) BS == 1)
```

Both AC1 and AC2u (for updated) hold.

```{r}
simpBottle$AC(\(BS) BS == 1, \(ST) ST == 1)
```

The reverse doesn't for AC2u.


## Complicated example from Fig 2.3 in the book


Now a more complicated example from the book (Fig 2.3).

```{r bottle}
bottle <- CausalMod$new(
  funs = list(
    ST = \() 1,
    BT = \() 1,
    SH = \(ST) ST,
    BH = \(BT, SH) BT * (1 - SH),
    BS = \(SH, BH) max(SH, BH)
  ), 
  ranges = list(
    ST = c(0, 1),
    BT = c(0, 1),
    SH = c(0, 1),
    BH = c(0, 1),
    BS = c(0, 1)    
  )
)
```

ST causes BT:

```{r}
bottle$AC(\(ST) ST == 1, \(BS) BS == 1)
```

Blimey. Imagine doing that by hand (though we didn't actually need to try all partitions and settings).

Reverse:

```{r}
bottle$AC(\(BS) BS == 1, \(ST) ST == 1)
```

It works!


## Two people throwing bricks at a glasshouse


Now a moderately hard one (not quite as bad as above) which I use in the [blog post](https://www.andifugard.info/actual-causes/).

```{r mod_hard}
mod_hard <- CausalMod$new(
  funs = list(
    AliceThrow  = \() 1,
    BobThrow    = \() 1,
    GlassBreaks = \(AliceThrow, BobThrow) max(AliceThrow, BobThrow)
  ), 
  ranges = list(
    AliceThrow  = c(0, 1),
    BobThrow    = c(0, 1),
    GlassBreaks = c(0, 1)
  )
)
```

```{r}
mod_hard$AC(\(AliceThrow) AliceThrow == 1, \(GlassBreaks) GlassBreaks == 1)
```


```{r}
mod_hard$AC(\(GlassBreaks) GlassBreaks == 1,
            \(AliceThrow) AliceThrow == 1)
```

## Scriven example


```{r scriven_ex}
scriven_ex <- CausalMod$new(
  funs = list(
    jumped_t0       = \() 1,
    shot_mid_flight = \() 1,
    dead_t1         = \(jumped_t0, shot_mid_flight)
                                min(jumped_t0, shot_mid_flight),
    dead_t2         = \(jumped_t0) jumped_t0
  ), 
  ranges = list(
    jumped_t0       = c(0, 1),
    shot_mid_flight = c(0, 1),
    dead_t1         = c(0, 1),
    dead_t2         = c(0, 1)
  )
)
```


Does jumping at t0 lead to death at t2?

```{r}
scriven_ex$AC(\(jumped_t0) jumped_t0 == 1,
              \(dead_t2)   dead_t2 == 1)
```


Does jumping at t0 cause death at t1?

```{r}
scriven_ex$AC(\(jumped_t0) jumped_t0 == 1,
              \(dead_t1)   dead_t1 == 1)
```


```{r}
scriven_ex$AC(\(shot_mid_flight) shot_mid_flight == 1,
              \(dead_t1) dead_t1 == 1)
```



```{r}
scriven_ex$AC(\(shot_mid_flight) shot_mid_flight == 1,
              \(dead_t2) dead_t2 == 1)
```