---
title: "Explore the impact of prior distributions in a simple single predictor Bayesian regression model"
author: "Andi Fugard ([@andi@sciences.social](https://sciences.social/@andi))"
date: "Last knitted `r format(Sys.Date(), '%d %B %Y')`"
output: 
  html_notebook: 
    code_folding: none
---


Includes required packages:

```{r}
library(conflicted)
library(brms)
library(tidyverse)
library(gtable)
library(tidybayes)
library(ggthemes)
library(tictoc)
library(beepr)
conflicts_prefer(
  brms::ar,
  stats::filter,
  stats::lag,
  brms::dstudent_t,
  brms::pstudent_t,
  brms::qstudent_t,
  brms::rstudent_t
)
```

Simulate a tiny dataset with n = 15.


```{r}
set.seed(1335)
the_n <- 15
dat <- tibble(x = rnorm(the_n, 0, 1),
              y = 2 * x + rnorm(the_n, 0, 10))

ggplot(dat, aes(x,y)) +
  geom_point()
```


```{r}
ols_tiny <- lm(y ~ x, data = dat)
summary(ols_tiny)
```


So the true slope for _x_ is 2, the OLS estimate from this sample is `r coef(ols_tiny)[2] |> round(2)`.

Here are the prior distributions on _x_ I want to try:

```{r}
priors <- expand.grid(mean = c(0, 2), sd = c(0.1, 1, 10)) |>
  mutate(prior = paste0("normal(", mean, ", ", sd, ")"))
priors
```

Fit the models:

```{r}
tic()
res <- map(
         priors$prior,
         \(p) brm(
                y ~ x,
                data = dat,
                prior = set_prior(p, class = "b", coef = "x"),
                silent = 0
              )
       )
toc()
beep(5)
```


Save draws from the posteriors of all these models into one data frame:

```{r}
names(res) <- priors$prior

grab_posteriors <- function(prior) {
  res[[prior]] |>
    spread_draws(b_x) |>
    mutate(prior = prior)
}

all_posteriors <- map(names(res), grab_posteriors) |> bind_rows()
```


Go again with a bigger dataset:

```{r}
set.seed(1335)
the_n <- 1500
dat_bigger <- tibble(x = rnorm(the_n, 0, 1),
                     y = 2 * x + rnorm(the_n, 0, 10))

ggplot(dat_bigger, aes(x, y)) +
  geom_point()
```

```{r}
lm(y ~ x, data = dat_bigger) |> summary()
```

```{r}
tic()
res_bigger <- map(
  priors$prior,
  \(p) brm(
         y ~ x,
         data = dat_bigger,
         prior = set_prior(p, class = "b", coef = "x"),
         silent = 0
       )
)
toc()
beep(3)
```



```{r}
names(res_bigger) <- priors$prior
grab_posteriors_bigger <- function(prior) {
  res_bigger[[prior]] |>
    spread_draws(b_x) |>
    mutate(prior = prior)
}
all_posteriors_bigger <- map(names(res_bigger), grab_posteriors_bigger) |> bind_rows()
```


Make some pictures:

```{r}
all_posteriors |>
  ggplot(aes(x = b_x)) +
  geom_histogram(bins = 75) +
  facet_wrap(vars(prior), scales = "free_y") +
  theme_few() +
  labs(x = ~ beta[x],
       y = "Count",
       title = "Impact of prior distribution on posterior (N = 15)")
```

```{r}
all_posteriors_bigger |>
  ggplot(aes(x = b_x)) +
  geom_histogram(bins = 75) +
  facet_wrap(vars(prior), scales = "free_y") +
  theme_few() +
  labs(x = ~ beta[x],
       y = "Count",
       title = "Impact of prior distribution on posterior (N = 1500)")
```


And all glued together:

```{r}
all_posteriors_15_1500 <- bind_rows(
  all_posteriors |> mutate(n = 15),
  all_posteriors_bigger |>  mutate(n = 1500)
)
```


```{r fig.height=3, fig.width=7, dpi=300}
final_plot <- all_posteriors_15_1500 |>
  ggplot(aes(x = b_x)) +
  geom_density() +
  facet_grid(cols = vars(prior), rows = vars(paste("n =", n)), scales = "free_y") +
  theme_few() +
  labs(x = ~ beta[x],
       y = "Density") +
  xlim(2 - 3, 2 + 3)
final_plot
```


```{r}
ggsave("final_plot_bayes.png", final_plot, height = 3, width = 7, dpi = 300)
```



Let's see if we can drag the posterior to cover 2 when the prior is N(0, 0.1), with a huge n = 1,000,000.


```{r}
set.seed(1335)
the_n <- 1e6
dat_biggest <- tibble(x = rnorm(the_n, 0, 1),
                      y = 2 * x + rnorm(the_n, 0, 10))

ggplot(dat_bigger, aes(x, y)) +
  geom_point()
```

```{r}
tic()
big_mod <- brm(
  y ~ x,
  data = dat_biggest,
  prior = set_prior("normal(0, 0.1)", class = "b", coef = "x"),
  silent = 0
)
toc()
beep(3)
```



```{r}
summary(big_mod)
```

```{r}
big_mod_pic <- big_mod |>
  spread_draws(b_x) |>
  ggplot(aes(x = b_x)) +
  geom_histogram(bins = 101) +
  theme_few() +
  labs(x = ~ beta[x], y = "Count") +
  xlim(2 - .1, 2 + .1)
big_mod_pic 
```

It's getting there -- you need a big sample size to undo a silly prior.


```{r}
ggsave("big_mod_pic.png", big_mod_pic, height = 3, width = 7, dpi = 300)
```



```{r}
lm(y ~ x, data = dat_biggest) |> confint()
```

