---
title: "Associations between gender and sexuality in the England and Wales 2021 Census"
author: "@Andi@lgbt.tech"
output:
  html_document: 
    theme: yeti
---

I really want to know whether your gender predicts your sexuality. ONS hasn't released the relevant crosstabs yet, so here's an approximation using variation in population counts across (lower-tier) local authorities. Beware the [ecological fallacy](https://en.wikipedia.org/wiki/Ecological_fallacy#Individual_and_aggregate_correlations), e.g., this might show that areas with more people of a particular sexuality also have more people of a particular gender, but not necessarily that they are the same people.


```{r message=FALSE}
library(tidyverse)
library(car)
```

## Read in the data

```{r}
sexuality <- read.csv("census2021sexuality.csv") |>
  rename(LA_code = Lower.Tier.Local.Authorities.Code,  
         LA_name = Lower.Tier.Local.Authorities,        
         sexuality_code = Sexual.orientation..9.categories..Code,
         sexuality = Sexual.orientation..9.categories.,
         n = Observation)
```

```{r}
head(sexuality)
```

```{r}
gender <- read.csv("census2021gender.csv") |>
  rename(LA_code = Lower.Tier.Local.Authorities.Code,  
         LA_name = Lower.Tier.Local.Authorities,        
         gender_code = Gender.identity..8.categories..Code,
         gender = Gender.identity..8.categories.,
         n = Observation)
```

```{r}
head(gender)
```

## Calculate per-milles


```{r}
sex_wide <- sexuality |>
  group_by(LA_code, LA_name) |>
  mutate(perc = n * 1000 / sum(n)) |>
  pivot_wider(
    id_cols = c(LA_code, LA_name),
    names_from = sexuality,
    values_from = perc
  )
```

```{r}
gender_wide <- gender |>
  group_by(LA_code, LA_name) |>
  mutate(perc = n * 1000 / sum(n)) |>
  pivot_wider(
    id_cols = c(LA_code, LA_name),
    names_from = gender,
    values_from = perc
  )
```

## Glue together

```{r}
sex_gender <- full_join(sex_wide, gender_wide, by = c("LA_code", "LA_name")) |>
  ungroup()
```

```{r}
head(sex_gender)
```

```{r}
names(sex_gender)
```

## OLS

```{r}
lm(
  `Non-binary` ~ `Straight or Heterosexual` +
    `Gay or Lesbian` +
    `Bisexual` + `Pansexual` + Queer + Asexual,
  data = sex_gender) |>
  confint() |> round(2)
```

LAs with more queer and asexual people have more nonbinary people.

```{r}
lm(
  `Trans man` ~ `Straight or Heterosexual` +
    `Gay or Lesbian` +
    `Bisexual` + `Pansexual` + Queer + Asexual,
  data = sex_gender) |>
  confint() |> round(2)
```

Areas with more pansexual people and fewer queer and asexual people have more trans men.

```{r}
lm(
  `Trans woman` ~ `Straight or Heterosexual` +
    `Gay or Lesbian` +
    `Bisexual` + `Pansexual` + Queer + Asexual,
  data = sex_gender) |>
  confint() |> round(2)
```

Areas with more pansexual people and fewer queer and asexual people also have more trans women.



## A picture

Let's try to get this in one diagram.


```{r message=FALSE}
library(MPsychoR)
library(qgraph)
```


```{r}
forcors <- sex_gender |>
  select(`Straight or Heterosexual`:`All other gender identities`) |>
  select(-c(ends_with(".x"), ends_with(".y"))) |>
  rename(
    Het = "Straight or Heterosexual",
    GL  = "Gay or Lesbian",
    Bi  = "Bisexual",
    Pan = "Pansexual",
    Ace = "Asexual",
    Q   = "Queer",
    Osx = "All other sexual orientations",
    Cis = "Gender identity the same as sex registered at birth",
    NCI = "Gender identity different from sex registered at birth but no specific identity given",
    TM  = "Trans man",
    TW  = "Trans woman",
    NBi = "Non-binary",
    Ogn = "All other gender identities"
  ) |>
  select(-c(Osx,NCI,Ogn))
```

```{r}
names(forcors)
```

```{r}
cors <- cor(forcors)
```


```{r dpi=600}
qgraph(
  cors,
  palette = "ggplot2",
  layout = "spring",
  graph = "glasso",
  sampleSize = nrow(sex_gender),
  threshold = TRUE
)
```

* Het = Straight or Heterosexual
* GL  = Gay or Lesbian
* Bi  = Bisexual
* Pan = Pansexual
* Ace = Asexual
* Q   = Queer
* Cis = Gender identity the same as sex registered at birth
* TM  = Trans man
* TW  = Trans woman
* NBi = Non-binary


