---
title: "Two ways to calculate the probability that if it rains, I have an umbrella"
author: Andi Fugard
date: 2 Nov 2024
output: 
  html_notebook: 
    code_folding: none
---

This post shows two ways to calculate the conditional probability that I have an umbrella, given that it's raining, $P(umbrella|raining)$; one using the usual ratio formula and the other using a three-valued logical connective, the conditional event, also known as the suppositional.

```{r}
library(conflicted)
library(tidyverse)
conflicts_prefer(dplyr::filter)
```

Here's a simulated dataset about rain and umbrellas, where each row is a day.

```{r}
days_n <- 20

set.seed(5)
sim_dat <- data.frame(day = 1:days_n, raining = rbinom(days_n, 1, .5)) |>
  mutate(umbrella = rbinom(days_n, 1, ifelse(raining, .8, .2))) |>
  mutate(across(raining:umbrella, as.logical))
sim_dat
```

The variable _raining_ denotes whether it was raining and _umbrella_ denotes whether I had an umbrella with me.

Here are the joint probabilities:

```{r}
joints <- sim_dat |>
  group_by(raining, umbrella) |>
  tally() |>
  ungroup() |>
  mutate(p = n / sum(n)) |>
  select(-n) |>
  arrange(desc(raining), desc(umbrella))
joints
```


What's the probability that if it rains, I have an umbrella, $P(umbrella | raining)$?

One way to do it is through:

$$P(umbrella|raining) = \frac{P(raining \land umbrella)}{P(raining)}$$

$P(raining \land umbrella)$ is:

```{r}
raining_and_umbrella_p <- joints |>
  filter(raining & umbrella) |>
  pull(p)
raining_and_umbrella_p 
```

$P(raining)$ is:

```{r}
raining_p <- joints |>
  filter(raining == 1) |>
  pull(p) |>
  sum()
raining_p
```

So, $P(umbrella|raining)$ is:

```{r}
raining_and_umbrella_p / raining_p
```

Another way to calculate this is using the _conditional event_, $B \dashv A$ ("B given A"), here using Hailperin's (1996) formulation (he calls it the _suppositional_):

```{r}
dont_care <- function(x) {
  ifelse(x, NA, x)
}

`%-|%` <- function(y, x) {
  dont_care(!x) | (x & y)
}
```

$B \dashv A$ is a connective in a three-valued logic with the values:

```{r}
vals <- c(TRUE, FALSE)
expand.grid(A = vals, B = vals) |>
  mutate(`B -| A` = B %-|% A) |>
  arrange(desc(A), desc(B))
```


We can apply this to the original dataset:

```{r}
sim_dat <- sim_dat |>
  mutate(`umbrella -| raining` = umbrella %-|% raining)
sim_dat
```


Now for $P(umbrella|raining)$ we drop the NAs and count the proportion of the time that $umbrella \dashv raining$ is TRUE:

```{r}
sim_dat |>
  pull(`umbrella -| raining`) |>
  mean(na.rm = TRUE)
```


## References

Hailperin, T. (1996). _Sentential Probability Logic. Origins, Development, Current Status, and Technical Applications_. Lehigh University Press.

