---
title: "CRAN package popularity"
author: "@Andi@tech.lgbt"
output: 
  html_notebook: 
    code_folding: none
---


```{r}
library(tidyverse)
library(dlstats)
library(ggrepel)
library(lubridate)
```


`cran_stats` gives us downloads per month, so they will be incomplete for the current month. I'll exclude the current month to account for this.

Here's a plot function, factored out of a previous copy and paste mess:

```{r}
plot_em <- function(dl_dat) {
  first_of_this_month <- floor_date(today(), unit = "month")
  
  dl_dat |>
  filter(start < first_of_this_month) |>
  mutate(label = if_else(start == max(start),
                         as.character(package),
                         NA_character_)
         ) |>
  ggplot(aes(
    x = start,
    y = downloads,
    colour = package,
    label = label
  )) +
  geom_smooth(se = FALSE) +
  geom_point() +
  labs(x = "Date", y = "Downloads", colour = "Package") +
  geom_label_repel(na.rm = TRUE, segment.colour = NA) +
  theme(legend.position = "none") +
  scale_x_date(expand = expansion(mult = 0.2)) +
  scale_y_continuous(labels = scales::comma)
}
```



(Some) mixed models and SEM packages:

```{r}
pack_stats <-
  cran_stats(c("lme4", "nlme", "sem", "lavaan", "blme", "brms", "rstanarm"))
```


```{r dpi = 300}
plot_em(pack_stats)
```

Same again the log scale:

```{r}
plot_em(pack_stats) +
scale_y_continuous(labels = scales::comma, trans = "log10")
```




Packages for matching and propensity score weighting:

```{r}
match_packages <-
  cran_stats(c(
    "Matching",
    "MatchIt",
    "twang",
    "PSweight",
    "cem"
  ))
```


```{r}
plot_em(match_packages)
```


```{r}
plot_em(match_packages) +
scale_y_continuous(labels = scales::comma, trans = "log10")
```


These packages are all installed as part of the tidyverse, but can also be installed individually.


```{r}
core_tidyverse_stats <-
  cran_stats(c(
    "ggplot2",
    "dplyr",
    "tidyr",
    "readr",
    "purrr",
    "tibble",
    "stringr",
    "forcats"
  ))
```

```{r}
plot_em(core_tidyverse_stats)
```



