---
title: "Playing with futures"
author: Andi Fugard
date: 13 July 2026
output: 
  html_notebook: 
    code_folding: none
---


I recently attended Henrik Bengtsson's excellent UseR! 2026 workshop on the {futureverse}. Here are a few examples to show what it does.


```{r}
library(future)
library(future.mirai)
library(tictoc)
library(tidyverse)
```


I'll use this to create a slow running process.

```{r}
rnorm_mean <- function(n, mean = 0, sd = 1) {
  rnorm(n, mean, sd) |> mean()
}
```


First go sequential:

```{r}
set.seed(42)
res_list <- list()

tic()
res_list$big1   <- rnorm_mean(2^27)
res_list$big2   <- rnorm_mean(2^27)
res_list$little <- rnorm_mean(20)
toc()
```

Here are the means:

```{r}
res_list
```

Now same again concurrently. I've added in a "huge" variant to demo something...


```{r}
plan(mirai_multisession, workers = 5)
set.seed(42)
future_list <- list()

tic()
future_list$big1   <- rnorm_mean(2^27) |> future(seed = TRUE)
future_list$big2   <- rnorm_mean(2^27) |> future(seed = TRUE)
future_list$huge   <- rnorm_mean(2^28) |> future(seed = TRUE)
future_list$little <- rnorm_mean(20)   |> future(seed = TRUE)
toc()
```

Those have zoomed off on concurrent processes so R is still able to run further commands.

```{r}
6 * 7
```

Check the value of the "little" one. This would block until it's done but it should have been swift, despite being run last.

```{r}
tic()
future_list$little |> value()
toc()
```

Now take a look at the two "big" ones -- this will block until they're done:

```{r}
tic()
future_list[c("big1", "big2")] |> map(value)
toc()
```

Take a look to see whether "huge" is done yet (and check the other completed ones, for completeness).

```{r}
future_list |> map(resolved)
```

Can't really be arsed waiting, so cancel it:

```{r}
cancel(future_list$huge)
```

Check out its status:

```{r}
future_list$huge
```

The real power of {future} is how it automagically futurizes calls like maps and bootstraps, hiding the *future* and *value* calls, but I found these examples handy to get to grips with what it's doing.


