---
title: "Encode bytes as hashtags"
author: "@andi@tech.lgbt"
output: 
  html_notebook: 
    code_folding: none
---

**Problem:** you want to encode arbitrary binary data as a hashtag on Mastodon.

**A solution:** transform to a hex string and shift the 0..f up to a..p so all the bytes are lowercase Latin characters.


First, load up {tidyverse}:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```


I want to transform a string like "Hello" to a hex string:

```{r}
chars2hexstr <- function(t) {
  t |>
    charToRaw() |>
    as.character() |>
    paste(collapse = "")
}
```

Testing:

```{r}
chars2hexstr("Hello")
```

Consulting an [ASCII table](https://en.wikipedia.org/wiki/ASCII#Printable_characters), looks good. "H" = 0x48, "e" = 0x65, "l" = 0x6c, and "o" = 0x6f.


Let's undo it:

```{r}
hexstr2chars <- function(hexstr) {
  substring(hexstr,
            seq(1, nchar(hexstr) - 1, 2),
            seq(2, nchar(hexstr), 2)) |>
  strtoi(base = 16) |>
  as.raw() |>
  rawToChar()
}
```


And a test:

```{r}
hexstr2chars("48656c6c6f")
```

Next, I want to shift 0..f up to a..p. I'm sure there's an easier way. Here's how I'm doing it. First, setup a mapping:

```{r}
charmap <- data.frame(
  hex = c(0:9, letters[1:6]),
  let = letters[1:16])
charmap
```

These two functions translate back and forth between 0..f (hex) and a..p (I'm going to call it a lethex), using a `strsplit` and `join`:

```{r}
hex2lethex <- function(hstr) {
  data.frame(hex = (hstr |> strsplit(""))[[1]]) |>
    left_join(charmap, by = "hex") |>
    pull(let) |>
    paste(collapse = "")
}

lethex2hex <- function(lethex) {
  data.frame(let = (lethex |> strsplit(""))[[1]]) |>
    left_join(charmap, by = "let") |>
    pull(hex) |>
    paste(collapse = "")
}
```

Try it with "Hello". Here's the hex string:

```{r}
hello_hex <- chars2hexstr("Hello")
hello_hex
```
Now as a lethex:

```{r}
hello_lethex <- hello_hex |> hex2lethex()
hello_lethex
```

And back again:

```{r}
hello_lethex |> lethex2hex()
```

Finally, a couple of wrappers, which take a string to a lethex string and back again:

```{r}
str2hashtag <- function(s) {
  s |> chars2hexstr() |> hex2lethex()
}

hashtag2str <- function(h) {
  h |> lethex2hex() |> hexstr2chars()
}
```


```{r}
str2hashtag("Hello")
```

```{r}
hashtag2str("eigfgmgmgp")
```

Finally:

```{r eval=FALSE}
hashtag2str("gihehehahddkcpcphhhhhhcohjgphfhehfgcgfcogdgpgncphhgbhegdgidphgdngefbhhdehhdjfhghfigdfb")
```
