---
title: "Playing with {rtoot}"
author: "@andi@tech.lgbt"
date: 13 Nov 2022
output: 
  html_notebook: 
    code_folding: none
---

I want to know if there's a positive correlation between the total number of users on servers and how many followers I have from those servers.

Include the packages I want...

```{r}
library(rtoot)
library(stringr)
library(tidyverse)
```


Get {rtoot} authorised to talk to my server:

```{r}
auth_setup(instance = "tech.lgbt", type ="user")
```

I'll need my ID: who am I?

```{r}
acc <- search_accounts("@Andi@tech.lgbt")
acc |>
  select(id, acct, display_name) |>
  head(1)
```

This is me!

```{r}
whoami <- "109273348690338129"
```


There's a handy function in {rtoot} for getting all followers; however, this doesn't currently support auto-pagination. After [reading the friendly manual](https://github.com/schochastics/rtoot/wiki/Pagination), here's a workaround:

```{r}
really_get_all_followers <- function(id, sure = "No!") {
  stopifnot(sure == "Yes, I know what I am doing")
  
  followers <- c()
  still_working <- TRUE
  max_id <- NULL
  
  while (still_working) {
    next_lot <- get_account_followers(id,
                                      max_id = max_id)
    followers <- bind_rows(followers, next_lot)
  
    attrs <- attr(next_lot, "headers")
    if ("max_id" %in% names(attrs))
      max_id <- attrs$max_id
    else
      still_working <- FALSE
  }
  
  followers
}
```


Get my followers:

```{r}
my_followers <- really_get_all_followers(
  whoami,
  sure = "Yes, I know what I am doing"
)
```


This number is correct: it worked!

```{r}
nrow(my_followers)
```


What server are they from?

```{r}
get_servers <- function(followers) {
  servers <- followers$acct |> str_split_fixed("@", 2)
  servers[,2]
}
```


```{r}
followers_servers <- data.frame(server = get_servers(my_followers)) |>
  mutate(server = ifelse(server == "", "tech.lgbt", server))
```

Here are the counts:

```{r}
server_count <- followers_servers |>
  group_by(server) |>
  summarise(n = n()) |>
  arrange(desc(n))

server_count
```

Check everything adds up:

```{r}
server_count$n |> sum()
```

So far so good.

Next up, how many users are there on each of those servers? Note the exception handling...

```{r}
get_user_count <- Vectorize(function(server) {
  cat("Trying: ")
  cat(server)
  cat("\n")
  res <- NA
  
  # This will catch problems like missing servers
  tryCatch(
    res <- get_instance_general(server)$stats$user_count,
    error = function(e) print(e)
  )
  res
})
```

```{r}
server_count$server_user_n <- get_user_count(server_count$server)
```


Finally, a graph:

```{r}
server_count |>
  na.omit() |>
  mutate(home = ifelse(server == "tech.lgbt",
                                 "Yes",
                                 "No"),
         `My home server` = factor(home,
                                   c("Yes", "No"))) |>
  ggplot(aes(y = log(n, 10),
             x = log(server_user_n, 10),
             colour = `My home server`)) +
  geom_point() +
  scale_colour_manual(values=c("purple", "black")) +
  ylab(expression(log[10]~(followers))) +
  xlab(expression(log[10]~(total~server~users))) +
  labs(title = "Follower count by server")
```


