---
title: "Mediator effects"
author: "Andi Fugard"
date: 30 June 2023
output: 
  html_notebook: 
    code_folding: none
---

```{r message=FALSE, warning=FALSE}
library(mediation)
```



Fit a model to estimate the total effect. *treat* is the treatment and *depress2* the outcome. The other variables are covariates.

```{r}
a <- lm(depress2 ~ treat + econ_hard + sex + age, data=jobs)
summary(a)
```

Now the models for the mediation test:

```{r}
b <- lm(job_seek ~ treat + econ_hard + sex + age, data=jobs)
c <- lm(depress2 ~ treat + job_seek + econ_hard + sex + age, data=jobs)
```


Estimate the CIs for the various effects by nonparametric bootstrap:

```{r}
contcont.boot <- mediate(b, c, boot=TRUE, sims=1000, treat="treat", mediator="job_seek")
```


```{r}
summary(contcont.boot)
```

Average direct effect (ADE):

```{r}
coef(c)["treat"]
```

Average causal mediation effect (ACME):

```{r}
coef(b)["treat"] * coef(c)["job_seek"]
```


Total effect:

```{r}
coef(a)["treat"]
```
Or the ADE + ACME:

```{r}
coef(c)["treat"] + coef(b)["treat"] * coef(c)["job_seek"]
```


Proportion mediated is ACME / total effect:

```{r}
coef(b)["treat"] * coef(c)["job_seek"] / coef(a)["treat"]
```



