---
title: "teffects ra, with bootstrapped SE, in R"
author: "Andi Fugard ([@andi@sciences.social](https://sciences.social/@andi))"
date: "Last knitted `r format(Sys.Date(), '%d %B %Y')`"
output:
  html_notebook:
    code_folding: none
  html_document:
    df_print: paged
---

```{r}
library(conflicted)
library(tidyverse)
library(haven)
library(boot)
library(tictoc)
library(beepr)
library(car)
library(marginaleffects)
```

```{r}
dat <- read_dta("https://www.stata-press.com/data/r14/cattaneo2.dta")
```


Try to replicate the example at:

https://www.stata.com/manuals/teteffectsra.pdf#teteffectsra

ATE is -239.6392, SE = 23.82402, z = -10.06, 95% CI = [-286.3334, -192.945].

ATT is -223.3017, SE = 22.7422, z = -9.82, 95% CI = [-267.8755, -178.7278].


## {marginaleffects}

Fit the model:

```{r}
gmod <- lm(bweight ~ mbsmoke * (prenatal1 + mmarried + mage + fbaby), data = dat)
```

ATE:

```{r}
avg_comparisons(gmod, variables = "mbsmoke", vcov = "HC1") |>
  print(digits = 7, p_eps = .001)
```

The estimate is spot on, the rest is very close, e.g., SE = 23.8086 vs 23.82402 (I think HC1 is the flavour of robust SE provided by Stata). Note the options passed to print to get more decimal places; the default gave me zero dp (the digits parameter is the number of significant figures).

Check ATT:

```{r}
avg_comparisons(
  gmod,
  variables = "mbsmoke",
  type = "response",
  newdata = subset(dat, mbsmoke == 1),
  vcov = "HC1"
) |>
  print(digits = 7, p_eps = .001)
```

Again, the estimate is identical and SE close: 22.54283 vs. Stata's 22.7422.



## Replicate teffects "by hand"

```{r}
boot_ate <- function(d, i) {
  the_dat <- d[i,]
  dat0 <- the_dat |> dplyr::filter(mbsmoke == 0)
  dat1 <- the_dat |> dplyr::filter(mbsmoke == 1)
  lm0 <- lm(bweight ~ prenatal1 + mmarried + mage + fbaby, data = dat0)
  lm1 <- lm(bweight ~ prenatal1 + mmarried + mage + fbaby, data = dat1)
  
  dat0 <- dat0 |>
    mutate(bweight_0 = bweight,
           bweight_1 = predict(lm1, newdata = dat0))
  
  dat1 <- dat1 |>
    mutate(bweight_1 = bweight,
           bweight_0 = predict(lm0, newdata = dat1))
  
  dat_all <- bind_rows(dat0, dat1) |>
    mutate(te = bweight_1 - bweight_0)
  
  res_ATE <- dat_all |>
    summarise(ate = mean(te))
  
  res_ATT <- dat_all |>
    dplyr::filter(mbsmoke == 1) |>
    summarise(att = mean(te))
  
  c(ate = res_ATE$ate, att = res_ATT$att)
}
```


I've packaged it all up a function ready for bootstrapping, which applied to the original data replicates teffects:

```{r}
boot_ate(dat, 1:nrow(dat))
```

Now bootstrap boot_ate to get SEs:

```{r}
tic()
boots <- boot(data = dat, statistic = boot_ate, R = 1499)
toc()
beep(2)
```


```{r}
res <- summary(boots) |> as.data.frame()
rownames(res) <- names(boots$t0)
res
```





