---
title: "Scrape books"
author: "Andi Fugard ([@andi@sciences.social](https://sciences.social/@andi))"
date: "Last updated 17 August 2024"
output:
  html_document:
    df_print: paged
---


```{r}
library(conflicted)
library(rvest)
library(kableExtra)
library(stringdist)
library(tidyverse)
```


Scrape, with a little help from clues I've added to the page:

```{r}
web_lines <- read_html("https://andifugard.info/misc/books/") |>
  html_text() |>
  read_lines()
dat <- tibble(raw = web_lines)

start_row <- which(dat$raw == "Note to self: start scraping here!") + 1
end_row   <- which(dat$raw == "Note to self: stop scraping here!")  - 1

scraped <- dat |>
  slice(start_row:end_row)
```

Separate author and title into a tibble:

```{r}
title_authormess <- str_split_fixed(scraped$raw, ", by | – ", n = 3)
my_books <- tibble(Author = title_authormess[, 2] |> trimws(),
                   Title  = title_authormess[, 1] |> trimws())

current_heading <- NA
my_books$Category <- rep(NA, nrow(my_books))
for (r in 1:nrow(my_books)) {
  if (my_books$Title[r] != "" && my_books$Author[r] == "")
    current_heading <- my_books$Title[r]
  
  if (my_books$Title[r] != "" && my_books$Author[r] != "")
    my_books$Category[r] <- current_heading  
}

my_books <- na.omit(my_books)
```



## Big list


Ta-da:

```{r}
my_books |>
  arrange(Category) |>
  kable(format = "html") |>
  kable_styling() |>
  column_spec(1, width = "10em") |>
  column_spec(2, width = "20em") |>
  column_spec(3, width = "5em")
```


## Fav authors

These are apparently my favourite authors (two books or more; todo: split coauthored books):

```{r}
my_books |>
  group_by(Author, Category) |>
  tally() |>
  arrange(desc(n)) |>
  arrange(Category) |>
  dplyr::filter(n > 1) |>
  kable(format = "html") |>
  kable_styling()
```



## Check for typos

I used this to look for mild one-off misspellings of authors' names mentioned twice or more:

```{r}
all_author_pairs <- my_books$Author |>
  unique() |>
  combn(m = 2) |>
  t()
colnames(all_author_pairs) <- c("a1", "a2")

all_author_pairs |>
  as_tibble() |>
  mutate(dist = stringdist(a1, a2)) |>
  arrange(dist) |>
  slice_head(n =10)
```


## Export to csv

I'll use this one day to export csv for a BookWyrm instance.

```{r}
var_names <- c("title", "author_text", "remote_id", "openlibrary_key", "inventaire_id", "librarything_key", "goodreads_key", "bnf_id", "viaf", "wikidata", "asin", "aasin", "isfdb", "isbn_10", "isbn_13", "oclc_number", "start_date", "finish_date", "stopped_date", "rating", "review_name", "review_cw", "review_content", "review_published", "shelf", "shelf_name", "shelf_date")
```




## Play around with APIs (to be continued)

```{r}
library(httr2)
library(jsonlite)
```


```{r}
lookup_isbn <- function(author, title) {
  base_url <- "https://www.googleapis.com/books/v1/volumes"
  query <- paste0("inauthor:",
                  URLencode(author),
                  "+intitle:",
                  URLencode(title))
  url <- paste0(base_url, "?q=", query)
  
  req <- request(url) %>%
    req_headers("Accept" = "application/json")
  
  resp <- req_perform(req)
  
  if (resp_status(resp) == 200) {
    book_data <- resp_body_json(resp)
    if (length(book_data$items) > 0) {
      isbn_list <- book_data$items[[1]]$volumeInfo$industryIdentifiers
      return(isbn_list)
    } else {
      return(NULL)
    }
  } else {
    return(NULL)
  }
}

lookup_info <- function(isbn) {
  url <- paste0("https://openlibrary.org/api/books?bibkeys=ISBN:", isbn, "&format=json&jscmd=data")
  req <- request(url) |>
    req_headers("Accept" = "application/json")
  
  resp <- req_perform(req)
  
  if (resp_status(resp) == 200) {
    book_data <- resp_body_json(resp)
    return(book_data)
  } else {
    return(NULL)
  }
}
```



```{r}
book_num <- 8

the_isbn <- lookup_isbn(my_books$Author[book_num], 
                        my_books$Title[book_num])[[1]]$identifier
book_info <- lookup_info(the_isbn)
book_info[[1]]$subjects
```


