---
title: "Scrape books"
author: "Andi Fugard ([@andi@sciences.social](https://sciences.social/@andi))"
date: "Last updated 17 August 2024"
output:
  html_document:
    df_print: paged
---


```{r}
library(conflicted)
library(rvest)
library(kableExtra)
library(stringdist)
library(tidyverse)
```


Scrape, with a little help from clues I've added to the page:

```{r}
web_lines <- read_html("https://andifugard.info/misc/books/") |>
  html_text() |>
  read_lines()
dat <- tibble(raw = web_lines)

start_row <- which(dat$raw == "Note to self: start scraping here!") + 1
end_row   <- which(dat$raw == "Note to self: stop scraping here!")  - 1

scraped <- dat |>
  slice(start_row:end_row)
```

Separate author and title into a tibble:

```{r}
title_authormess <- str_split_fixed(scraped$raw, ", by | – ", n = 3)
my_books <- tibble(Author = title_authormess[, 2] |> trimws(),
                   Title  = title_authormess[, 1] |> trimws())

current_heading <- NA
my_books$Category <- rep(NA, nrow(my_books))
for (r in 1:nrow(my_books)) {
  if (my_books$Title[r] != "" && my_books$Author[r] == "")
    current_heading <- my_books$Title[r]
  
  if (my_books$Title[r] != "" && my_books$Author[r] != "")
    my_books$Category[r] <- current_heading  
}

my_books <- na.omit(my_books)
```



## Big list


Ta-da:

```{r}
my_books |>
  arrange(Category) |>
  kable(format = "html") |>
  kable_styling() |>
  column_spec(1, width = "10em") |>
  column_spec(2, width = "20em") |>
  column_spec(3, width = "5em")
```


## Fav authors

First split out coauthored books into single authors:

```{r}
authors_mat <- str_split_fixed(my_books$Author, "&|,|,&", n = Inf)
colnames(authors_mat) <- paste0("Author_", 1:ncol(authors_mat))

authors_tib <- as_tibble(authors_mat)

my_books_coauth <- bind_cols(my_books, authors_tib) |>
  dplyr::select(-Author) |>
  pivot_longer(cols = starts_with("Author_"),
               values_to = "Author") |>
  dplyr::filter(Author != "") |>
  mutate(Author = trimws(Author))
```


These are apparently my favourite authors (two books or more):

```{r}
my_books_coauth |>
  group_by(Author) |>
  tally() |>
  arrange(desc(n), Author) |>
  dplyr::filter(n > 1) |>
  kable(format = "html") |>
  kable_styling()
```

```{r}
fav_authors <- my_books_coauth |>
  group_by(Author) |>
  tally() |>
  arrange(desc(n), Author) |>
  dplyr::filter(n > 1) |>
  pull(Author)

titles_of_authors <- my_books_coauth |>
  dplyr::filter(Author %in% fav_authors) |>
  pull(Title)
```

Books by these authors:

```{r}
my_books |>
  dplyr::filter(Title %in% titles_of_authors) |>
  dplyr::arrange(Author, Title) |>
  kable(format = "html") |>
  kable_styling() |>
  column_spec(1, width = "10em") |>
  column_spec(2, width = "20em") |>
  column_spec(3, width = "5em")
```





## Check for typos

I used this to look for mild one-off misspellings of authors' names mentioned twice or more:

```{r}
all_author_pairs <- my_books_coauth$Author |>
  unique() |>
  combn(m = 2) |>
  t()
colnames(all_author_pairs) <- c("a1", "a2")

all_author_pairs |>
  as_tibble() |>
  mutate(dist = stringdist(a1, a2)) |>
  arrange(dist) |>
  slice_head(n =10)
```


## Export to csv


For importing into a BookWyrm instance:

```{r}
for_csv <- my_books_coauth |>
  group_by(Title) |>
  summarise(author = paste(Author, collapse = ", ")) |>
  rename(title = Title) |>
  mutate(shelf = "read")
#write_csv(for_csv, "books.csv")
```
