---
title: "Examples from Senn's LinkedIn post, The Main Chance"
author: Andi Fugard
date: 13 Aug 2024
output:
  html_document:
    df_print: paged
---



```{r}
library(conflicted)
library(tidyverse)
library(tidymodels)
```

## Datasets to explore

This one is from https://www.linkedin.com/pulse/main-chance-stephen-senn-c0u2e/


```{r}
senn_dat <- read.csv(text = "treat,sex,events,patients
Ctr,F,18,100
Int,F,5,100
Ctr,M,5,50
Int,M,4,50")  |>
  mutate(sex = as.factor(sex), treat = as.factor(treat))
senn_dat
```



```{r}
no_interaction <- read.csv(text = "treat,sex,events,patients
Ctr,F,20,100
Int,F,4,100
Ctr,M,10,50
Int,M,2,50") |>
  mutate(sex = as.factor(sex), treat = as.factor(treat))
no_interaction
```


```{r}
strong_interaction <- read.csv(text = "treat,sex,events,patients
Ctr,F,20,100
Int,F,4,100
Ctr,M,2,50
Int,M,10,50") |>
  mutate(sex = as.factor(sex), treat = as.factor(treat))
strong_interaction
```


## Select one of the datasets

```{r}
dat <- senn_dat
```



## Explore models

Statistician A:

```{r}
modA <- glm(cbind(events, patients - events) ~ treat * sex,
            data = dat,
            family = binomial)
summary(modA)
```


Statistician B:

```{r}
modB <- glm(
  cbind(events, patients - events) ~ treat * relevel(sex, "M"),
  data = dat,
  family = binomial
)
summary(modB)
```


Statistician C:

```{r}
dat$sexC <- dat$sex
contrasts(dat$sexC) <- c(-1, 1)
contrasts(dat$sexC)
```


```{r}
modC <- glm(
  cbind(events, patients - events) ~ treat * sexC,
  data = dat,
  family = binomial
)
summary(modC)
```

Statistician D:

```{r}
modD_M <- glm(
  cbind(events, patients - events) ~ treat,
  data = dat,
  family = binomial,
  subset = sex == "M"
)
modD_F <- glm(
  cbind(events, patients - events) ~ treat,
  data = dat,
  family = binomial,
  subset = sex == "F"
)
```

```{r}
res <- bind_rows(tidy(modD_M) |> mutate(sex = "M"),
                 tidy(modD_F) |> mutate(sex = "F")) |>
  dplyr::filter(term == "treatInt") |>
  mutate(var    = std.error ^ 2,
         weight = 1 / var) |>
  dplyr::select(-c(statistic, p.value))

res
```

Pooled effect, SE, and z

```{r}
pooled_mod <- tibble(
  term = "treatInt",
  estimate = sum(res$estimate * res$weight) / sum(res$weight),
  std.error = sqrt(1/sum(res$weight)),
  statistic = estimate / std.error,
  p.value = pnorm(abs(statistic), lower.tail = FALSE) * 2
)

pooled_mod
```


## Compare all the estimates:

```{r}
bind_rows(
tidy(modA) |> mutate(mod = "A"),
tidy(modB) |> mutate(mod = "B"),
tidy(modC) |> mutate(mod = "C"),
pooled_mod |> mutate(mod = "D")
) |>
  dplyr::filter(term == "treatInt") |>
  select(mod, estimate, std.error, statistic, p.value) |>
  mutate(across(where(is.numeric), \(x) round(x,3)))
```



## Another approach -- check the interaction

Check the interaction for model A, B, or C -- all give the same answer:

```{r}
bind_rows(
  drop1(modA, test = "Chi") |> tidy() |> mutate(mod = "A"),
  drop1(modB, test = "Chi") |> tidy() |> mutate(mod = "B"),
  drop1(modC, test = "Chi") |> tidy() |> mutate(mod = "C")
) |>
  dplyr::filter(term != "<none>") |>
  dplyr::select(mod, df, LRT, p.value)
```


When looking only at main effects of treatment, it doesn't matter how sex is coded:

```{r}
modG1 <- glm(
  cbind(events, patients - events) ~ treat + relevel(sex, "F"),
  data = dat,
  family = binomial
)
modG2 <- glm(
  cbind(events, patients - events) ~ treat + relevel(sex, "M"),
  data = dat,
  family = binomial
)

modG3 <- glm(
  cbind(events, patients - events) ~ treat + sexC,
  data = dat,
  family = binomial
)
```

```{r}
bind_rows(modG1 |> tidy() |> mutate(mod = "G1"),
          modG2 |> tidy() |> mutate(mod = "G2"),
          modG3 |> tidy() |> mutate(mod = "G3")) |>
  mutate(term = ifelse(str_detect(term, "sex"), "sex", term)) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  select(mod, term, estimate, std.error) |>
  arrange(term)
```









