---
title: "Simulating efficacy trials"
author: Andi Fugard
date: 12 May 2024
output: 
  html_notebook: 
    code_folding: none
---


```{r}
library(conflicted)
library(tidyverse)
library(tictoc)
```


```{r include=FALSE}
tic()
```


Let's simulate some data where we know the average treatment effect.

```{r}
set.seed(3137942)
theN <- 500

sim_dat <- tibble(latent_magic = 50 + rnorm(theN, 0, sqrt(.6)),
                  y0 = latent_magic + rnorm(theN, 0, sqrt(.4)),
                  y1 = latent_magic + 0.2 + rnorm(theN, 0, sqrt(.4)),
                  ITS = y1 - y0) |>
  select(-latent_magic)
```

Take a look:

```{r}
sim_dat |>
  slice_head(n = 10) |>
  mutate(across(everything(), \(x) round(x, 1)))
```

y0 is the outcome under control, y1 the outcome under treatment, and ITS is the individual treatment effect, y1 -- y0.

We can get the true average treatment effect of an efficacy trial by averaging all those ITSes:


```{r}
trueATE <- mean(sim_dat$ITS)
trueATE
```

That's the true answer we're trying to recover with an RCT.


Now let's simulate an RCT.

```{r}
rct_dat <- sim_dat |>
  mutate(
    treat = rbinom(n(), 1, .5),
    y = treat * y1 + (1 - treat) * y0,
    y1 = ifelse(treat == 1, y1, NA),
    y0 = ifelse(treat == 0, y0, NA)
  ) |>
  select(-ITS)

rct_dat |>
  slice_head(n = 10) |>
  mutate(across(everything(), \(x) round(x, 1)))
```

We only see one of the potential outcomes, depending on which was realised.

Estimate ATE and its 95% CI:

```{r}
lm_mod <- lm(y ~ treat, data = rct_dat)
cbind(ATE = coef(lm_mod), confint(lm_mod))["treat",]
```

Repeat that 2,000 times to see the average of the ATEs across simulated trials:


```{r}
sim_reses <- data.frame()

for (i in 1:2000) {
  rct_dat <- sim_dat |>
    mutate(treat = rbinom(n(), 1, .5),
           y = treat * y1 + (1 - treat) * y0)
  
  lm_mod <- lm(y ~ treat, data = rct_dat)
  this_res <- cbind(ATE = coef(lm_mod), confint(lm_mod))["treat", ]
  
  sim_reses <- bind_rows(sim_reses, this_res)
}
```


```{r}
sim_reses  |>
  slice_head(n = 10) |>
  mutate(across(everything(), \(x) round(x, 2)))
```


The mean estimated ATE across simulated trials is *really* close to the true ATE...


```{r}
c(
  `mean of ATE estimates` = mean(sim_reses$ATE),
  `true ATE` = trueATE,
  `true ATE - mean of estimates` = mean(sim_reses$ATE) - trueATE
) |> round(4)
```


... but there's quite a bit of variation around the mean answer.


```{r}
sim_reses |>
  ggplot(aes(x = ATE)) +
  geom_histogram(bins = 40) +
  xlab("estimated ATE")
```


Let's look at the 95% confidence interval coverage:

```{r}
sim_reses <- sim_reses |>
  mutate(trueATE_in_interval = trueATE >= `2.5 %` &
                               trueATE <= `97.5 %`)
cover <- mean(sim_reses$trueATE_in_interval)
cover
```


`r round(cover * 100,0)`% of the simulated studies' 95% confidence intervals included the true result.


```{r echo=FALSE}
toc()
```


