---
title: 'Surreal numbers: work in progress'
author: "Andi Fugard"
output:
  html_document:
    df_print: paged
---

Here are my [notes and links to useful resources on surreals](https://www.andifugard.info/surreal-numbers/).

Please don't use this to do any sums for safety critical systems ;-)

This is a helper function:

```{r is_integer}
is_integer <- \(x) x == as.integer(x)
```

```{r}
is_integer(42)
```

```{r}
is_integer(pi)
```

I represent surreals as lists. This is another helper to make them and apply the class "surreal".

```{r as.surreal}
as.surreal <- function(l) {
  res <- l
  
  if(is.null(res) || length(l) == 0)
    res <- nullset
  
  class(res) <- "surreal"
  res
}
```

Rather than sets of surreals, I'm using lists and ignoring duplicates and order in the output...

```{r}
as.surreal_list <- function(sl) {
  res <- sl
  class(res) <- "surreal_list"
  res
}
```

Some useful predicates:

```{r is-surreal-surreals}
is.surreal <- function(s)
  class(s) == "surreal"

is.surreal_list <- function(sl)
  class(sl) == "surreal_list"

is.nullset <- function(s) {
  is.surreal(s) && length(s) == 0
}
```

And the main way to construct surreals and lists of surreals:

```{r surreal-surreals}
surreal <- function(l, r) {
  list(l = l,
       r = r) |> as.surreal()
}

surreals <- function(...) {
  res <- list(...) |> unique()
  class(res) <- "surreal_list"
  res
}
```

How to peek into the left or right pair of a surreal. Note that if there's nothing there to peek at, I just return an empty set.

```{r LR}
L <- function(s) {
  stopifnot(class(s) == "surreal")
  
  if (is.null(s) || length(s) == 0)
    nullset
  else
    s$l
}

R <- function(s) {
  stopifnot(class(s) == "surreal")
  
  if (is.null(s) || length(s) == 0)
    nullset
  else
    s$r
}
```

And the empty set. Well, list.

```{r nullset}
nullset <- list()
class(nullset) <- "surreal"
```

This seems to work... At present it just removes duplicates and nullsets. It could also go through and remove elements in L that are smaller than the largest surreal and elements of R that are larger than the smallest surreal.

```{r}
simplify_surreals <- function(s) {
  if (is.surreal(s)) {
    if (is.nullset(s))
      nullset
    else
      surreal(simplify_surreals(s |> L()),
              simplify_surreals(s |> R()))
  }
  else if (is.surreal_list(s)) {
    if (length(s) == 0)
      nullset
    else if (length(s) == 1)
      simplify_surreals(s[[1]])
    else {
      # find nulls and zap them
      res <- c()
      for (one_s in s) {
        one_res <- simplify_surreals(one_s)
        if (!is.nullset(one_res))
          res <- c(res, list(one_res))
      }
      res |> unique() |> as.surreal_list()
    }
  }
  else
    stop("No idea what to do with that, sorry")
}
```

This simplifies dyadic numbers, $n/2^k$.

```{r simplify_dyadic}
simplify_dyadic <- \(n, k) {
  stopifnot(is_integer(n))
  stopifnot(is_integer(k))
  stopifnot(k >= 0)
  
  while (k > 0 && n %% 2 == 0) {
    n <- n/2
    k <- k - 1
  }
  
  c(n, k)
}
```

This is Tøndering's "dali" function that builds a surreal from a dyadic number.

```{r dali}
# x = n/2^k
dali <- \(n, k, loud = FALSE) {
  stopifnot(is_integer(n))
  stopifnot(is_integer(k))
  stopifnot(k >= 0)
  
  if (loud) {
    cat("dali called with ")
    cat(n); cat("/"); cat(2^k)
  }
  
  n_old <- n
  k_old <- k
  
  simple <- simplify_dyadic(n,k)
  n <- simple[1]
  k <- simple[2]
  
  if (loud && (n != n_old || k != k_old)) {
    cat(" = "); cat(n); cat("/"); cat(2^k)
  }
  
  if (loud) {
    cat(" = "); cat(n/2^k); cat("\n")
  }
  
  res <- NA
  
  if (n == 0) {
    res <- list(l = nullset, r = nullset)
  }
  else if (k == 0) {
    if (n > 0)
      res <- list(l = dali(n - 1, 0), r = nullset)
    else
      res <- list(l = nullset, r = dali(n + 1, 0))      
  }
  else {
    res <- list(l = dali(n-1, k), r = dali(n+1, k))
  }
  
  class(res) <- "surreal"
  res
}
```

Print surreals:

```{r print.surreal}
print.surreal <- function(s) {
  stopifnot(class(s) == "surreal")
  stopifnot(length(s) %in% c(0,2))
  
  if (length(s) == 0) {
    cat("∅")
  }
  else if ((length(s) == 2) &&
           all(names(s) == c("l", "r"))) {
    cat("{")
    print(s |> L())
    cat(" | ")
    print(s |> R())
    cat("}")    
  }
  else
    stop("I don't know how to print that surreal")
}
```

And lists of surreals:

```{r print.surreal_list}
print.surreal_list <- function(surs) {
  stopifnot(class(surs) == "surreal_list")
  
  if (length(surs) == 0)
    cat("∅")
  else {
    i <- 1
    for (s in surs) {
      print(s)
      if (i < length(surs))
        cat(", ")
      i <- i+1
    }
  }
}
```

(In)equalities:

```{r}
less_equal <- function(x, y) {
  greater_equal(y, x)
}

greater_equal <- function(x, y) {
  if (is.nullset(x) || is.nullset(y))
    FALSE # BEWARE: I don't know why this isn't TRUE...
          # but it seems to work as is 
  else {
    res <- c() 

    if (is.surreal(x |> R()))
      res <- c(res, less_equal(x |> R(), y))
    else
      for (x_R in x |> R())
        res <- c(res, less_equal(x_R, y))
    
    if (is.surreal(y |> L()))
      res <- c(res, less_equal(x, y |> L()))
    else    
      for (y_L in y |> L())
        res <- c(res, less_equal(x, y_L))

    !any(res)
  }
}
```

```{r}
equal <- function(x,y)
  greater_equal(x,y) && less_equal(x,y)

less <- function(x,y)
  less_equal(x,y) && !equal(x,y)

greater <- function(x,y)
  greater_equal(x,y) && !equal(x,y)
```

Given all the above, addition is relatively painless. You will see I haven't done multiplication yet.

```{r add}
add <- function(a, b) {
  if (length(a) == 0 || length(b) == 0) {
    nullset
  }
  else {
    surreal(l = surreals(add(a |> L(), b),
                         add(a, b |> L())),
            r = surreals(add(a |> R(), b),
                         add(a, b |> R())))
  }
}
```

And some examples:

```{r}
add(dali(0,0), dali(0,0))
```

```{r}
add(dali(1,0), dali(1,0)) |> simplify_surreals()
```

```{r}
add(dali(1,0), dali(1,1))
```

```{r}
add(dali(1,0), dali(1,1)) |> simplify_surreals()
```


```{r}
dali(1,0) # 1
```
```{r}
dali(2,0) # 2
```
```{r}
# 1+1 = 2
add(dali(1,0), dali(1,0)) |> simplify_surreals()
```

Hurrah, so 1+1 = 2. Great.

```{r}
add(dali(2,0), dali(-2,0)) |> simplify_surreals()
```

That mess above is equal to zero.

```{r}
equal(add(dali(2,0), dali(-2,0)), dali(0,0))
```

```{r}
raw_sum <- add(dali(2,2), dali(2,2))
simp_sum <- raw_sum |> simplify_surreals()
correct_ans <- dali(1,0)
```

```{r}
raw_sum
```


```{r}
simp_sum
```

```{r}
correct_ans
```


```{r}
equal(raw_sum, simp_sum)
```

```{r}
equal(simp_sum, correct_ans)
```

Let's check whether a putative surreal really is:

```{r}
legit_surreal <- function(s) {
  !greater_equal(s |> L(), s |> R())
}
```

YES:

```{r}
legit_surreal(dali(0,1))
```

NOPE:

```{r}
legit_surreal(surreal(dali(0,1), dali(0,1)))
```