---
title: "Subtraction on integers encoded as natural numbers"
author: Andi Fugard
output:
  html_document:
    df_print: paged
---

```{r message=FALSE}
library(tidyverse)
library(broom)
```


You can represent all infinitely many integers ($\mathbb{Z}$),

..., -3, -2, -1, 0, 1, 2, 3, ...

using natural numbers (\mathbb{N}*)

0, 1, 2, 3, ...

by mapping nonnegative numbers, $z$, to even numbers $2z$ and all negative numbers to odd numbers $-2z - 1$:

```{r}
PZ <- function(z) {
  ifelse(z >= 0, 2*z, -2*z - 1)
}
```

Here are some examples:

```{r}
tibble(z  = -3:3,
       pz = PZ(z))
```

And to reverse:

```{r}
Z <- function(pz) {
  ifelse(pz %% 2 == 0, pz/2, -(pz + 1)/2)
}
```

Seems to work...

```{r}
-3:3 %>% PZ() %>% Z()
```

Now, how might we define subtraction on these representations of $\mathbb{Z}$, assuming that out here at the meta-level we can use negatives but in the representation it's only ever possible to save a natural number. I will assume that we have addition, multiplication, and subtraction on naturals.

Some helper functions:

```{r}
even <- \(x) x %% 2 == 0
odd  <- \(x) !even(x)
```


And a bunch of examples:

```{r}
examples <- expand.grid(x = -5:5, y = -5:5) %>%
  mutate(
    x_pz     = PZ(x),
    y_pz     = PZ(y),
    x_abs_pz = PZ(abs(x)), # I'm guessing these
    y_abs_pz = PZ(abs(y)), # will be handy
    
    # Some key conditions that will help us do the sums
    c1       = ifelse(even(x_pz), "even(x_pz)", "odd(x_pz)"),
    c2       = ifelse(even(y_pz), "even(y_pz)", "odd(y_pz)"),
    c3       = case_when(x_abs_pz < y_abs_pz ~ "x_abs_pz < y_abs_pz",
                         x_abs_pz > y_abs_pz ~ "x_abs_pz > y_abs_pz",
                         TRUE                ~ "x_abs_pz == y_abs_pz"),
    
    # Paste the conditions together as an R conjunction:
    cond     = paste(c1, c2, c3, sep = " & "),
    
    # Now do subtraction and work out what it's like represented in N:
    r        = x - y,
    r_pz     = PZ(x - y)
  )
```


There are 12 combinations of conditions:

```{r}
examples$cond %>% unique()
```



```{r}
slice_sample(examples, n = 5)
```


Next up, I'm going to use... wait for it... linear regression... to estimate the answer for each of these 12 cases:

```{r}
mods <- examples %>%
  nest(data = -cond) %>% 
  mutate(
    fit = map(data, ~ lm(r_pz ~ x_pz * y_pz, data = .x)),
    tidied = map(fit, tidy)
  ) %>% 
  unnest(tidied)
mods
```

Transform the answer into an output suitable for a `case_when`:

```{r}
mods_wide <- mods %>%
  select(cond, term, estimate) %>%
  mutate(estimate = ifelse(is.na(estimate),
                           0,
                           round(estimate, 5))) %>%
  pivot_wider(id_cols = "cond",
              names_from = "term",
              values_from = "estimate") %>%
  arrange(`(Intercept)`, x_pz, y_pz) %>%
  mutate(
    formula = paste0(`(Intercept)`, " + ",
                     x_pz, " * x_pz + ",
                     y_pz, " * y_pz "),
    case = paste("    ", cond, "~\n      ", formula)
  )
```

For ease of copy and paste:

```{r}
cat(mods_wide$case, sep = ",\n")
```


This function was mostly written by pasting in the answer above:

```{r}
pz_sub <- function(x_pz, y_pz) {
  x_abs_pz <- ifelse(even(x_pz), x_pz, x_pz + 1)
  y_abs_pz <- ifelse(even(y_pz), y_pz, y_pz + 1)
  
  case_when(
     even(x_pz) & even(y_pz) & x_abs_pz < y_abs_pz ~
       -1 + -1 * x_pz + 1 * y_pz ,
     odd(x_pz) & odd(y_pz) & x_abs_pz > y_abs_pz ~
       -1 + 1 * x_pz + -1 * y_pz ,
     odd(x_pz) & odd(y_pz) & x_abs_pz < y_abs_pz ~
       0 + -1 * x_pz + 1 * y_pz ,
     odd(x_pz) & odd(y_pz) & x_abs_pz == y_abs_pz ~
       0 + 0 * x_pz + 0 * y_pz ,
     even(x_pz) & even(y_pz) & x_abs_pz == y_abs_pz ~
       0 + 0 * x_pz + 0 * y_pz ,
     even(x_pz) & even(y_pz) & x_abs_pz > y_abs_pz ~
       0 + 1 * x_pz + -1 * y_pz ,
     odd(x_pz) & even(y_pz) & x_abs_pz > y_abs_pz ~
       0 + 1 * x_pz + 1 * y_pz ,
     odd(x_pz) & even(y_pz) & x_abs_pz < y_abs_pz ~
       0 + 1 * x_pz + 1 * y_pz ,
     even(x_pz) & odd(y_pz) & x_abs_pz == y_abs_pz ~
       0 + 2 * x_pz + 0 * y_pz ,
     even(x_pz) & odd(y_pz) & x_abs_pz < y_abs_pz ~
       1 + 1 * x_pz + 1 * y_pz ,
     even(x_pz) & odd(y_pz) & x_abs_pz > y_abs_pz ~
       1 + 1 * x_pz + 1 * y_pz ,
     odd(x_pz) & even(y_pz) & x_abs_pz == y_abs_pz ~
       1 + 2 * x_pz + 0 * y_pz
  )
}
```


Give it a test:

```{r}
examples <- examples %>% mutate(
  r_pz_est = pz_sub(x_pz, y_pz)
)
```


Did it work...?

```{r}
examples %>%
  select(x, y, r, x_pz, y_pz, r_pz, r_pz_est)
```


Yes.

```{r}
examples %>%
  filter(r_pz != r_pz_est)
```




